Transport map accelerated adaptive importance sampling, and application to inverse problems arising from multiscale stochastic reaction networks

### 2020-01-14 Resubmission/TETAIS.tex
First draft of resubmission complete

Other approaches to challenging posterior structure have been
proposed, including the work of Marzouk and Parno\cite{parno2014transport}, who
demonstrated how invertible transport maps can be constructed which
map the target distribution close to an easily-explored reference
density, such as a $d$-dimensional Gaussian. Since the map is
invertible, Gaussian proposal densities on the reference space can be
mapped to highly informed moves on the parameter space. A related idea
is the use of the anisotropic diffusion maps\cite{singer2008non}, in
which a map which describes the manifold is constructed using
principle component analysis. These ideas have also been applied to
stochastic reaction
networks\cite{singer2009detecting,dsilva2016data}. In
\cite{chiavazzo2017intrinsic}, these maps were used to propagate
data-mining into unexplored regions which were predicted to be close
to the lower-dimensional manifold. The idea is similar to that of the
Lamperti or Girsanov transformations for SDEs\cite{flandrin2003stationarity,gardiner2009stochastic},  the aim being to map complex
processes onto a standard one which is well understood, through a
change of measure.

@@

****Could move this material to stochastic network section****

@@

********Did I mean to get rid of this?**********

In this paper, we aim to incorporate transport maps into an ensemble
importance sampling setting, in order to develop stable algorithms which are
capable of sampling from probability distributions with complex
structures. These maps are constructed using an importance sample of
the posterior distribution, and are used to simplify the posterior
distribution, mapping them as close as possible to a reference
Gaussian measure. Through this map, straightforward Gaussian proposal
densities are transformed into complex densities, which match as well
as possible the target density. Less well-informed proposal kernels means that complex
structures in the density can only be well represented with a large
increase in the ensemble size, increasing computational cost. If
computational resources do not allow for such an increase, this can
lead to stability issues in the algorithm, due to the proposal
distribution in the importance sampler not representing well the
target density, leading to high variance in the weights, causing slow
convergence. Better informed proposal densities
leads to more stable and faster converging importance sampling
schemes, and we will demonstrate this through examples
arising from Bayesian inverse problems on multiscale stochastic
reaction networks.

@@

In importance sampling, we aim to sample from a target $\pi$, by
sampling from a chosen proposal distribution $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
sampling from a different distribution to $\pi$. Monte Carlo estimates
using a sample of size $N$
of a function $f$ with respect to $\pi$ can then be made through the
formula
\begin{equation} \mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
  w_kf(\theta^{(k)}),\end{equation}
where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
This method works well when $\pi$ and $\chi$ are close, but can be
excruciatingly slow when they are not. The idea behind adaptive
importance sampling methods is
to construct a good proposal distribution, either from the entire
history of the algorithm up to the current point, or to use the
current state of a whole ensemble of $M$ particles in the system.

@@

@@

In this paper, we investigate the use of transport maps to learn local
covariances across the whole of the domain, in order to stabilise
adaptive importance sampling methods, and make these methods more applicable to a wider
range of more challenging inference problems.

\section{Construction of transport maps in importance sampling} \label{sec:map}
%Follow the other papers, show adaptation(s)
In this Section, we describe the construction of transport maps which
allow for the simplification of complex posterior distributions in
order to allow for improved sampling, in particular for methods based
on importance sampling.
In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
distribution to the posterior distribution, the idea being that one could draw a moderately sized
sample from the prior distribution and use this sample to approximate a map onto the target space.
Once this map was known to the desired accuracy, a larger sample from the prior could be used to
investigate the posterior distribution. This
methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
distribution, the map transforms a sample from the posterior onto a reference space.
The reference density is chosen to allow efficient proposals using a simple proposal
distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.

@@

In this Section we outline the methodology in
\cite{parno2014transport} for approximately coupling the target
measure with density
$\pi_{\theta}$, with the reference distribution with density $\pi_r$, and
show how the map can be constructed using a weighted sample
and hence how we can incorporate the map into importance sampling schemes.

@@

Armed with such a map, independent samples can be made of the target
measure, using the pullback of the reference density $\phi$ through $T^{-1}$.
Clearly the pullback only exists when $T$ is monotonic, i.e. has a positive definite Jacobian, and has continuous first derivatives.
Not all maps satisfy these conditions, so we define a smaller space of
maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all
invertible maps. This space does not necessarily contain an exact
coupling between target and reference space, and so we are motivated to formulate an optimisation problem to find the map $\tilde{T}
\in \mathcal{T}^\uparrow$ which most closely maps the target density
to the reference density.

% We attempt to find the deterministic coupling of two continuous
% probability distributions, $(\mu_\theta, \hat{\mu_r})$, such that
% $\hat{\mu_r} = T\mu_\theta$ where the distance between $\mu_r$ (the
% desired reference measure) and $\hat{\mu_r}$ (the achieved reference
% measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
% \cite{parno2014transport} we aim to minimise the Kullback-Liebler (KL)
% divergence between the density of $\mu_{\theta}$ and the pullback of
% the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and
% $\tilde{\pi}(\theta)$. For two absolutely continuous measures with
% densities $\pi_1$ and $\pi_2$ respectively, the KL
% divergence is given by
% \begin{equation}D_\text{KL}(\pi_1\|\pi_2) = 
% 		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\end{equation}
% The KL divergence is not itself a norm, since
% it is not symmetric, i.e. $D_\text{KL}(\pi_1\|\pi_2) \neq
% D_\text{KL}(\pi_2\|\pi_1)$ in general. However, it is still a useful
% measure of the similarity of two probability distributions, not least since the
% square root of the KL-divergence is an upperbound to the Hellinger
% distance metric.

As in previous work in \cite{parno2014transport}, we can ensure
invertibility if we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
\begin{equation}
	\tilde{T}(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
\end{equation}
where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. % We assume that the target and reference probability densities are absolutely continuous on
% $\mathbb{R}^d$. Under this formulation, with appropriate
% regularisation (which will be detailed later), we are guaranteed a unique invertible map $\tilde{T}$ with the property that
% \begin{equation}
% 	\tilde{T}\mu_{\theta} \approx \mu_r.
% \end{equation}
% % Relaxing the equality constraint to finding the approximate map
% % $\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises $C(T)$, gives us a
% % practical route to finding a good candidate map.

\subsection{The optimisation problem}
Our aim is now to find the lower triangular map $\tilde{T} \in
\mathcal{T}^{\ltri}$ such that the difference between target
density and the pullback of the reference density is minimised. As in
\cite{parno2014transport}, we choose the cost function
to be the Kullback-Leibler (KL) divergence between the posterior density and the pullback density,
\begin{equation}
	D_\text{KL}(\pi\|\tilde{\pi}) =
		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
\end{equation}
This divergence results in some nice properties which we will explore in the following derivation. The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to its relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
\begin{equation}
	D_{KL}(p\|q) \geq D_H^2(p,q),
\end{equation}
which is a true distance metric on probability distributions.
Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
\begin{equation}
	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
		\log\left|J_{\tilde{T}}(\theta)\right|\right].
\end{equation}
We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function. This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown. However this is exactly the scenario for which we would turn to MCMC methods for a solution.

@@

The expectation in \eqref{eq:gen_map_optim}} can be
approximated by using a Monte Carlo approximation. The form of the penalisation term promotes maps which are
closer to the identity, and so prevents overfitting when the quality
or size of the current sample from the posterior is not sufficient.

\subsection{The structure of the map}*****CHECK HOW MUCH OF THIS IS
IDENTICAL TO MARZOUK PAPER******

Before we continue with the derivation of the optimisation problem, we consider the structure
of the map in more detail. The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$. The Jacobian of $\tilde{T}$ is a lower triangular matrix,
\begin{equation}
	J_T(\theta) = \begin{bmatrix}
		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
	\end{bmatrix} = \begin{bmatrix}
		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & 0\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
	\end{bmatrix}
\end{equation}
since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$. This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
\begin{equation}\label{eqn:separable_jacobian}
	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta),
\end{equation}
where we note that this term is separable in terms of the dimension $i$.

Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map. The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$. This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$. We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.

We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions. Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not necessarily orthogonal with respect to the target distribution. Each component of the map is defined as a multivariate polynomial expansion,
\begin{equation}\label{eq:map_defn}
	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
\end{equation}
The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
\mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
\begin{equation}
	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
\end{equation}
and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$. Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.

The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
optimisation problem, and so we would like to keep this number as small as possible. One option is
to use polynomials of total order $p$,
\begin{equation}
	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
\end{equation}
which is optimal in terms of the amount of information captured by the map about the target. The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = \binom{i+p}{p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$. Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$. These index sets are discussed
in~\cite{parno2014transport}. Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}. Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the ETAIS algorithm.

@@

We now discuss how we can evaluate the cost function in
Equation~\eqref{eq:gen_map_optim}. In \cite{parno2014transport}, this
expectation is approximated using an MCMC estimator, such that
\begin{align}
	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
			\beta\mathbb{E}(T(\theta)-\theta)^2 \notag \\
		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. 
\end{align}
Here we diverge from previous work, as we aim to build a map from
samples from an importance sampling scheme. Such samples no longer
carry equal weight, and as such the Monte Carlo estimator becomes
\begin{equation}\label{eqn:TM_full_cost}
	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
		w_k \left[-\log\pi_r(T_i(\theta^{(k)})) -
			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
\end{equation}
where $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
all these weights. Optimisation of this cost function results in a map from $\pi$ to some reference density $\pi_r$. By choosing the reference density to be a Gaussian density, we can simplify this expression greatly. Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
\begin{equation}\label{eqn:TETAIS_objective}
	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
\end{equation}

@@

These simplifications result in the efficiently implementable, regularised optimisation problem for
computing the map coefficients, 
\begin{align}\label{eq:weighted_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
%where additionally we have introduced a factor of $\frac{1}{K}$ to the
%regularisation term to prevent over-smoothing as the number of samples increases.
This optimisation problem can be efficiently solved using Newton iterations. It is suggested
in~\cite{parno2014transport} that this method usually converges in around 10-15 iterations, and we
have seen no evidence that this is not a reasonable estimate. When calculating the map several times
during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
criteria.

@@

Given importance samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
target measure to a reference measure. We now consider how to
implement an ensemble importance sampling algorithm which uses
these maps to propose new states. In \cite{parno2014transport} it was
shown how approximate transport maps can be used to accelerate
Metropolis-Hastings methods, with the map being periodically updated
with the samples produced from the target measure. Convergence of this
adaptation is shown in~\cite{parno2014transport}. In this Section, we
will show how similarly, these maps can be used to construct highly
efficient importance sampling schemes.

>>>>

Other approaches to challenging posterior structure have been
proposed, including the work of Marzouk and Parno\cite{parno2018transport}, who
demonstrated how invertible transport maps can be constructed which
map the target distribution close to an easily-explored reference
density, such as a $d$-dimensional Gaussian. Since the map is
invertible, Gaussian proposal densities on the reference space can be
mapped to highly informed moves on the parameter space. A related idea
is the use of the anisotropic diffusion maps\cite{singer2008non}, in
which a map which describes the manifold is constructed using
principle component analysis. These ideas have also been applied to
stochastic reaction
networks\cite{singer2009detecting,dsilva2016data}. In
\cite{chiavazzo2017intrinsic}, these maps were used to propagate
data-mining into unexplored regions which were predicted to be close
to the lower-dimensional manifold. The idea is similar to that of the
Lamperti or Girsanov transformations for SDEs\cite{flandrin2003stationarity,gardiner2009stochastic},  the aim being to map complex
processes onto a standard one which is well understood, through a
change of measure.

@@

% ****Could move this material to stochastic network section****

@@

% ********Did I mean to reintroduce this last paragraph?********** NOOOOO!!!

In this paper, we aim to incorporate transport maps into an ensemble
importance sampling setting, in order to develop stable algorithms which are
capable of sampling from probability distributions with complex
structures. These maps are constructed using an importance sample of
the posterior distribution, and are used to simplify the posterior
distribution, mapping them as close as possible to a reference
Gaussian measure. Through this map, straightforward Gaussian proposal
densities are transformed into complex densities, which match as well
as possible the target density. Less well-informed proposal kernels means that complex
structures in the density can only be well represented with a large
increase in the ensemble size, increasing computational cost. If
computational resources do not allow for such an increase, this can
lead to stability issues in the algorithm, due to the proposal
distribution in the importance sampler not representing well the
target density, leading to high variance in the weights, causing slow
convergence. Better informed proposal densities
leads to more stable and faster converging importance sampling
schemes, and we will demonstrate this through examples
arising from Bayesian inverse problems on multiscale stochastic
reaction networks. \edit{In a perfect scenario, the numerical
  approximation of the optimal transport map would capture all of the
  posterior structure, and independent samples could be drawn by
  sampling from the reference measure. However in many cases the push
  forward of the posterior measure can still exhibit complex structure
  which means that efficient adaptive importance sampling can be
  preferential to this independent sampling approach.}

@@

In importance sampling, we aim to sample from a target $\pi$, by
sampling from a chosen proposal distribution $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
sampling from a different distribution to $\pi$. Monte Carlo estimates
using a sample of size $N$
of a function $f$ with respect to $\pi$ can then be made through the
formula
\begin{equation} \mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
  w_kf(\theta^{(k)}),\end{equation}
where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
This method works well when $\pi$ and $\chi$ are close, but can be
excruciatingly slow when they are not. The idea behind adaptive
importance sampling methods is
to construct a good proposal distribution, either from the entire
history of the algorithm up to the current point, or to use the
current state of a whole ensemble of $M$ particles in the system.  \edit{A nice illustrative figure of how adaptive
  importance sampling algorithms work is given in \cite{bugallo2017adaptive}.}

@@

\edit{The choice of proposal kernel $q$ can affect the convergence of
  the algorithm. In particular, it is important that the tails of the
  proposal distribution are as fat as the target distribution. In this
  work we will primarily use Gaussian kernels, but we have previously
  made comparisons with more complex kernels, such as MALA or
  HMC\cite{cotter2015parallel,russ2017parallel}. There is minimal
  benefit from using these more complex kernels, which come at
  additional cost.}

@@

\edit{In this paper, we investigate the use of transport maps which allows
us to learn the local correlation structures of the posterior in
currently explored regions. Using these transport maps, we can
stabilise adaptive importance sampling methods, and make these methods
more applicable to a wider range of more challenging inference problems.}

\section{Construction of transport maps in importance sampling} \label{sec:map}
%Follow the other papers, show adaptation(s)
In this Section, we describe the construction of transport maps which
allow for the simplification of complex posterior distributions in
order to allow for improved sampling, in particular for methods based
on importance sampling.
In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
distribution to the posterior distribution, the idea being that one could draw a moderately sized
sample from the prior distribution and use this sample to approximate a map onto the target space.
Once this map was known to the desired accuracy, a larger sample from the prior could be used to
investigate the posterior distribution. This
methodology was adapted in~\cite{parno2018transport} to form a new proposal method for MH
algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
distribution, the map transforms a sample from the posterior onto a reference space.
The reference density is chosen to allow efficient proposals using a simple proposal
distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.

@@

In this Section we outline the methodology in
\cite{parno2018transport} for approximately coupling the target
measure with density
$\pi_{\theta}$, with the reference distribution with density $\pi_r$, and
show how the map can be constructed using a weighted sample
and hence how we can incorporate the map into importance sampling schemes.

@@

Armed with such a map, independent samples can be made of the target
measure, using the pullback of the reference density $\phi$ through $T^{-1}$.
Clearly the pullback only exists when $T$ is monotonic, i.e. has a positive definite Jacobian, and has continuous first derivatives.
Not all maps satisfy these conditions, so we define a smaller space of
maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$, \edit{ where $\mathcal{T}$ which contains all
invertible maps. It is an intractable problem to find the
  Knothe–Rosenblatt rearrangement, and so we numerically approximate
  it by formulating an optimisation problem to find the map $\tilde{T}
\in \mathcal{T}^\uparrow$ which most closely maps the target density
to the reference density.}

% We attempt to find the deterministic coupling of two continuous
% probability distributions, $(\mu_\theta, \hat{\mu_r})$, such that
% $\hat{\mu_r} = T\mu_\theta$ where the distance between $\mu_r$ (the
% desired reference measure) and $\hat{\mu_r}$ (the achieved reference
% measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
% \cite{parno2018transport} we aim to minimise the Kullback-Liebler (KL)
% divergence between the density of $\mu_{\theta}$ and the pullback of
% the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and
% $\tilde{\pi}(\theta)$. For two absolutely continuous measures with
% densities $\pi_1$ and $\pi_2$ respectively, the KL
% divergence is given by
% \begin{equation}D_\text{KL}(\pi_1\|\pi_2) = 
% 		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\end{equation}
% The KL divergence is not itself a norm, since
% it is not symmetric, i.e. $D_\text{KL}(\pi_1\|\pi_2) \neq
% D_\text{KL}(\pi_2\|\pi_1)$ in general. However, it is still a useful
% measure of the similarity of two probability distributions, not least since the
% square root of the KL-divergence is an upperbound to the Hellinger
% distance metric.

As in previous work in \cite{parno2018transport}, we can ensure
invertibility if we restrict the map to be \edit{lower triangular\cite{bogachev2005triangular}}, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. \edit{There are a number of different
  approaches to guaranteeing monotonicity\cite{y.17:_handb_uncer_quant}, but here we follow
  the approach taken in \cite{parno2018transport}.} This lower triangular map has the form,
\begin{equation}
	\tilde{T}(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
\end{equation}
where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. % We assume that the target and reference probability densities are absolutely continuous on
% $\mathbb{R}^d$. Under this formulation, with appropriate
% regularisation (which will be detailed later), we are guaranteed a unique invertible map $\tilde{T}$ with the property that
% \begin{equation}
% 	\tilde{T}\mu_{\theta} \approx \mu_r.
% \end{equation}
% % Relaxing the equality constraint to finding the approximate map
% % $\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises $C(T)$, gives us a
% % practical route to finding a good candidate map.

\subsection{The optimisation problem}
Our aim is now to find the lower triangular map $\tilde{T} \in
\mathcal{T}^{\ltri}$ such that the difference between target
density and the pullback of the reference density is minimised. As in
\cite{parno2018transport}, we choose the cost function
to be the Kullback-Leibler (KL) divergence between the posterior density and the pullback density,
\begin{equation}
	D_\text{KL}(\pi\|\tilde{\pi}) =
		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
\end{equation}
This divergence results in some nice properties which we will explore in the following derivation. The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to its relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
\begin{equation}
	D_{KL}(p\|q) \geq D_H^2(p,q),
\end{equation}
which is a true distance metric on probability distributions.
Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
\begin{equation}
	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
		\log\left|J_{\tilde{T}}(\theta)\right|\right].
\end{equation}
We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function. This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown. However this is exactly the scenario for which we would turn to MCMC methods for a solution.

@@

The expectation in \eqref{eq:gen_map_optim}} can be
approximated by using a Monte Carlo approximation. The form of the penalisation term promotes maps which are
closer to the identity, and so prevents overfitting when the quality
or size of the current sample from the posterior is not
sufficient. \edit{Moreover, we can exploit the lower triangular
  structure, for example in the evaluation of 
\begin{equation}\label{eqn:separable_jacobian}
	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta),
\end{equation}
where we note that this term is separable in terms of the dimension
$i$. Similarly, inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map.

As in \cite{parno2018transport}, we parameterise each component of the
map $\tilde{T}_i$ with a multivariate polynomial expansion,
\begin{equation}\label{eq:map_defn}
	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
\end{equation}
built from
a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not
necessarily orthogonal with respect to the target distribution. The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
\mathcal{J}_i \subset \mathbb{N}_0^i$. 
Here we stick with polynomials of total order $p$ since we work with
low dimensional problems with the ETAIS algorithm. More details about
the structure of the maps can be found in  \cite{parno2018transport}.}

% Each component of the map is defined as a multivariate polynomial expansion,
% \begin{equation}\label{eq:map_defn}
% 	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
% \gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
% \end{equation}
% The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
% $\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
% \mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
% \begin{equation}
% 	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
% \end{equation}
% and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$. Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.

% The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
% optimisation problem, and so we would like to keep this number as small as possible. One option is
% to use polynomials of total order $p$,
% \begin{equation}
% 	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
% \end{equation}
% which is optimal in terms of the amount of information captured by the map about the target. The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = \binom{i+p}{p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$. Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$. These index sets are discussed
% in~\cite{parno2018transport}. Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}.

@@

We now discuss how we can evaluate the cost function in
Equation~\eqref{eq:gen_map_optim}. In \cite{parno2018transport}, this
expectation is approximated using an MCMC estimator, such that
\begin{align}
	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
			\beta\mathbb{E}(T(\theta)-\theta)^2 \notag \\
		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. 
\end{align}
Here we diverge from previous work, as we aim to build a map from
samples from an importance sampling scheme. Such samples no longer
carry equal weight, and as such the Monte Carlo estimator becomes
\begin{equation}\label{eqn:TM_full_cost}
	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
		w_k \left[-\log\pi_r(T_i(\theta^{(k)})) -
			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
\end{equation}
where $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
all these weights. Optimisation of this cost function results in a map from $\pi$ to some reference density $\pi_r$. By choosing the reference density to be a Gaussian density, we can simplify this expression greatly. Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
\begin{equation}\label{eqn:TETAIS_objective}
	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
\end{equation}

@@

These simplifications result in the efficiently implementable, regularised optimisation problem for
computing the map coefficients, 
\begin{align}\label{eq:weighted_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
%where additionally we have introduced a factor of $\frac{1}{K}$ to the
%regularisation term to prevent over-smoothing as the number of samples increases.
This optimisation problem can be efficiently solved using Newton iterations. It is suggested
in~\cite{parno2018transport} that this method usually converges in around 10-15 iterations, and we
have seen no evidence that this is not a reasonable estimate. When calculating the map several times
during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
criteria.

@@

Given importance samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
target measure to a reference measure. We now consider how to
implement an ensemble importance sampling algorithm which uses
these maps to propose new states. In \cite{parno2018transport} it was
shown how approximate transport maps can be used to accelerate
Metropolis-Hastings methods, with the map being periodically updated
with the samples produced from the target measure. Convergence of this
adaptation is shown in~\cite{parno2018transport}. In this Section, we
will show how similarly, these maps can be used to construct highly
efficient importance sampling schemes.
